### 2022-03-10 bools/bools.tex
more

\begin{abstract}
    We present a pragmatic approach to extending a Boolean-free higher-order
    superposition calculus to support Boolean reasoning. Our approach extends
    inference rules that have been used only in a first-order setting, uses some
    well-known rules previously implemented in higher-order provers, as well as new rules.
    We have implemented the approach in the Zipperposition
    theorem prover. The evaluation shows highly competitive performance of our approach
    and clear improvement over previous techniques.
\end{abstract}

@@

In Chapter~\ref{ch:intro} we motivated our work by the need to bridge the gap
between higher-order frontends and first-order backends. This gap is
traditionally bridged using well-known translations from higher-order to
first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
\ref{ch:ehoh}, translations are usually less efficient than native support. The
distinguishing features of higher-order logic used by proof assistants that the
translation must eliminate include $\lambda$-binders, function
extensionality---the property that functions are equal if they agree on every
argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
\, z \ieq y \,z) \iimplies x \ieq y$, and formulas occurring as arguments of
function symbols \cite{mp-08-trans}.

\looseness=-1
A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate
$\lambda$-binders and functional extensionality. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on calculus-level
treatment of Booleans and the other, which requires no changes to the calculus,
based on preprocessing.

\looseness=-1 To fully bridge the gap between higher-order and first-order tools,
we combine the two approaches: we use the efficient higher-order superposition
calculus and extend it with inference rules that reason with
Boolean terms. In early work, Kotelnikov et al.\ 
\cite{kotelnikov-15-fool} have described a \emph{FOOL paramodulation}
rule that, under some order requirements, removes the need for the axiom
describing the Boolean domain---$\iforall p. \; p \ieq \itrue \llor p
\ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
arguments of symbols is translated to first-order logic.

@@

\looseness=-1 An important feature that we inherit by building on top of Bentkamp
et al. \cite{bbtvw-21-sup-lam} is support for (function) extensionality. Moving to higher-order
logic with Booleans also means that we need to consider \emph{Boolean extensionality}: $\iforall pq. \, (p
\iequiv q) \iimplies p \ieq q$. We extend the rules of Bentkamp et al.\ 
that treat function extensionality to also treat Boolean extensionality.

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the
higher-order provers Leo-III \cite{sb-21-leo3} and Satallax
\cite{cb-12-satallax}, such as elimination of Leibniz equality, primitive
instantiation and treatment of choice operator \cite{pa-01-classical-ty-thy}. We
have also designed new rules that use higher-order unification to resolve
Boolean formulas that are hoisted to literal level, delay clausification of nonatomic
literals, reason about formulas under $\lambda$-binders, and many others. Even
though the rules we use are inspired by the ones of refutationally complete
higher-order provers, we do not guarantee completeness of our extension of
$\lambda$-superposition.

@@

Our approach is implemented in the Zipperposition theorem prover
\cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
to implement their higher-order superposition calculus. We further extend
their implementation.

\looseness=-1
We performed an extensive evaluation of our approach (Sect.~\ref{sect:bool:eval}).
In addition to evaluating different configurations of our new rules, we have
compared them to full higher-order provers CVC4, Leo-III,  Satallax and Vampire.
The results suggest that it is beneficial to natively support Boolean reasoning
-- the approach outperforms preprocessing-based approaches. Furthermore, it is
very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
approaches we base on, as well as related approaches (Sect.~
\ref{sect:bool:discussion}).

@@

We base our work (and parts of the following text) on Bentkamp et al.'s \cite{bbtvw-21-sup-lam} extensional
polymorphic clausal higher-order logic. We extend the syntax of this logic by
adding logical connectives to the signature. The semantic of the logic
is extended by interpreting Boolean type $o$ as a two-element domain. This
amounts to extending Bentkamp et al's fragment of higher-order logic to
full-higher order logic (HOL). Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$ abstraction, polymorphic
types, and native logical connectives. For reference, we provide precise definition of this polymorphic logic.

@@

\looseness=-1
Function symbols $\cst{a}, \cst{b}, \cst{f}, \cst{g}, \ldots$ are elements of
$\Sigma$; their type declarations are written as $\cst{f} :
\forallty{\tuplen{\alpha}} \tau$. Set $\VV$ contains both free and bound variables, following
the distinction introduced in Sect.~\ref{sec:pre:hol}. Free term variables from the set $\VV$ are written
$F,G,X,Y \ldots$ and we denote their types as $X : \tau$. Bound term variables are written $x,y,z,\ldots$,
and their types are similarly denoted. When the type is not
important, we omit type declarations. We assume that symbols $\itrue,
\ifalse,\inot,\iand, \ior,\iimplies,\iequiv$ with their standard meanings and type declarations are elements of
$\Sigma$. Furthermore, we assume that polymorphic symbols $\iforall$ and $\iexists$
with type declarations $\forallty{\alpha} (\alpha \rightarrow o) \rightarrow o$
and $\ieq \; : \forallty{\alpha} {\alpha \rightarrow \alpha \rightarrow o}$ are
in $\Sigma$, with their standard meanings. All these symbols are called \emph{logical
symbols}. We use infix notation for binary logical symbols.

@@

\looseness=-1
Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q})
\llor \inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while
$f|_{1.2.1}$ is a nested Boolean, as well as its subterms. First-order logic
allows only top-level Booleans, whereas nested Booleans are characteristic for
higher-order logic.

@@

\looseness=-1
Zipperposition already had some support for Boolean reasoning before we
started extending the calculus of Bentkamp et al. In this section, we first
describe the internals of Zipperposition responsible for reasoning with
Booleans. We continue by describing \NumberOK{15} rules that we have
implemented. For ease of presentation we divide them in three categories.
\subsection{Support for Booleans in Zipperposition}
\label{subsect:bool:zip-bools}

@@

In Zipperposition, applications are represented in flattened, spine notation. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. 
Zipperposition's support for $\lambda$-terms is used to represent quantified nested
Booleans: Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of the
input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

@@

\looseness=-1
Kotelnikov et al.\ \cite{kotelnikov-15-fool}, to the best of our
knowledge, pioneered the approach of extending a first-order superposition prover to support
nested Booleans. They call effects of including the axiom $\iforall
p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
explosive behavior of the axiom, they imposed the following two requirements to
the simplification order $\succ$: $\itrue \succ \ifalse$ and
$\itrue$ and $\ifalse$ are two smallest ground terms with respect to $\succ$. If
these requirements are met, there is no self-paramodulation of the clause
and only paramodulation possible is from literal $p \eq \itrue$ of the mentioned axiom
into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
the axiom with the inference rule \emph{FOOL Paramodulation} (\infname{FP}):
\pagebreak[2]
%
$$ \namedinference{FP}{C[f]}{C[\itrue] \vee f \eq \ifalse} $$
%
\looseness=-1
where $f$ is a nested non-variable Boolean subterm of clause $C$, different from
$\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
Booleans  to first-order logic without interpreted Booleans; thus, symbols $\itrue$ and $\ifalse$, and type $o$
correspond to proxy symbols and types introduced during the translation.

We created two rules that are syntactically similar to \infname{FP} but
are adapted for higher-order logic with one key distinction---we do not perform any translation:
%
\begin{align*}
  & \namedinference{Cases}{C[f]}{C[\ifalse] \vee f \eq \itrue}
  && \namedsimp{CasesSimp}{C[f]}{C[\ifalse] \vee f \eq \itrue \quad C[\itrue] \vee f \noteq \itrue}
\end{align*}
%
\looseness=-1
Clearly, the prover that uses the rules should not include them both at the same time.
In addition, since
literals $f \eq \ifalse$ are represented as negative equations $f \noteq \itrue$, which cannot be used to paramodulate from,
we change the first requirement on the order to $\ifalse \succ
\itrue$.

\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
These two rules hoist Boolean subterms $f$ to the literal level; therefore,
some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
called immediate clausification (\infname{IC}):
%
$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
%
We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
is applicable if clause $C = L_1 \llor
\cdots \llor L_n$ is not standard.
The resulting clauses $\tuple{D}{m}$ represent
the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
Using Boolean extensionality, Zipperposition's clausification
algorithm treats Boolean equality as equivalence (i.e., it replaces
$\mathop{\eq}\typeargs{o}$ with $\iequiv$).

@@

\looseness=-1
If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result in many redundant clauses. Consider the following
example: let $\cst{p} : o \rightarrow o$, $\cst{a} : o$ and consider a clause
set containing $\cst{p} \, (\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq
\itrue$. Then, the clause $C = \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
derived in eight ways using the rules, depending on which nested Boolean subterm was
chosen for the inference. In general, if a clause has a subterm occurrence of the form $\cst{p}^n \, \cst{a}$,
where both $\cst{p}$ and $\cst{a}$ have result type $o$, the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in $2^{n-1}$ ways.
% Boolean subterms occurring in different
% eligible literals cause similar explosive behavior.
To combat these issues we
implemented pragmatic restrictions of the rule: only $f$ which is
the leftmost outermost (or innermost) eligible subterm will be considered. With
this modification $C$ can be derived in only one way. Furthermore,
some intermediate conclusions of the rules will not be derived, pruning the search space.

The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
eagerly simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
the formula $\cst p\land\itrue$ is replaced by $\cst p$. To simplify nested Booleans we implemented the rule

\[
\namedsimp{BoolSimp}{C[\sigmaterm{f}]}{C[\sigmaterm{g}]}
\]
where $f \longrightarrow g \in E$ runs over fixed set of rewrite rules $E$,
and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
described by Nonnengart and Weidenbach \cite[Sect.~3]{nw-01-small-cnf}. This set contains
the rules describing how each logical symbol behaves when given arguments $\itrue$ or $\ifalse$: for example, 
it includes $(\itrue \iimplies p)~\longrightarrow~p $ and $(p \iimplies \itrue) \longrightarrow \itrue$. %full list visa report Sect 4.2
Leo-III implements a similar rule, called \textsf{simp} \cite[Sect.~4.2.1.]{as-18-phd}.

@@

\subsection{Higher-Order Considerations}
\label{subsect:bool:core}
\looseness=-1
To achieve refutational completeness of higher-order resolution and similar
calculi it is necessary to instantiate variables with result type $o$,
\emph{predicate variables}, with arbitrary formulas
\cite{as-18-phd,pa-01-classical-ty-thy}. Fortunately, we can approximate the
formulas using a complete set of logical symbols (e.g., $\inot$, $\iforall$, and
$\iand$). Since such an approximation is not only necessary for completeness of
some calculi, but very useful in practice, we implemented the \emph{primitive
instantiation} (PI) rule:
%
$$ \namedinference{PI} {C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq}
\, t} {\substcl{\{ P
\mapsto f \}}{C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq} \, t}  } $$
%
where $P$ is a free variable of
the type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow o$. 
Choosing a different $f$ that instantiates $P$, we can balance between
explosiveness of approximating a complete set of logical symbols and
incompleteness of pragmatic approaches. We borrow the notion of imitation from
higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}), and we say
that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
\, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $Z$ of type $\nu_1
\rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
are fresh free variables, where each $Y_i$ has the type $\nu_1 \rightarrow
\cdots \rightarrow \nu_m \rightarrow \tau_i$; variable $x_i$ is of type $\nu_i$.
\pagebreak[2]

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. The rule has different modes that generate
sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
$\star$ is an element of a set of logical constants $P = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
contains imitations (for $P$) of all elements of $P$. Mode $\emph{Pragmatic}$
contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\ineq j$ and  $\tau_i = \tau_j$, it contains $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
\tau_j = o$, then it contains $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
contains $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ contains
imitations of $\itrue$, $\ifalse$ and $\star$ (except for $\mathit{Imit_{\iforall\iexists}}$ which contains imitations
of both $\iforall$ and $\iexists$).

\looseness=-1
While experimenting with our implementation we have noticed some proof patterns
that led us to come up with the following modifications. First, it often
suffices to perform \infname{PI} only on initial clauses---which is why we
allow the rule to be applied only to the clauses created using at most $k$
generating inferences. Second, if the rule was used in the proof, its premise
is usually only used as part of that inference---which is why we implemented
a version of \infname{PI} that removes the clause after all possible \infname{PI}
inferences have been performed. We observed that the mode \emph{Imit}$_\star$ is useful in practice since often only a single
approximation of a logical symbol is necessary.

>>>>

\begin{abstract}
    We present a pragmatic approach to extending a Boolean-free higher-order
    superposition calculus to support Boolean reasoning. Our approach extends
    inference rules that have been used only in a first-order setting, uses some
    rules previously implemented in higher-order provers, as well as new rules.
    We have implemented the approach in the Zipperposition
    theorem prover. The evaluation shows highly competitive performance of our approach
    and clear improvement over previous techniques.
\end{abstract}

@@

In Chapter~\ref{ch:intro} we motivated our work by the need to bridge the gap
between higher-order frontends and first-order backends. This gap is
traditionally bridged using translations from higher-order to
first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
\ref{ch:ehoh}, translations are usually less efficient than native support. The
distinguishing features of higher-order logic used by proof assistants that the
translation must eliminate include $\lambda$ binders, function
extensionality---the property that functions are equal if they agree on every
argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
\, z \ieq y \,z) \iimplies x \ieq y$---and formulas occurring as arguments of
function symbols \cite{mp-08-trans}.

\looseness=-1

% A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
% extension of superposition for extensional Boolean-free higher-order logic. The
% extension removes the need to translate $\lambda$-binders and functional
% extensionality. 
As mentioned in Sect.~\ref{sec:pre:ho-sup-calculi}, Bentkamp et al. developed
\lsup{}, a complete higher-order calculus that alleviates the need to translate
$\lambda$-terms and function extensionality. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on calculus-level
treatment of Booleans and the other, which requires no changes to the calculus,
based on preprocessing.

\looseness=-1 To fully bridge the gap between higher-order and first-order tools,
we combine the two approaches: We take \lsup{} as our basis and  extend it with inference rules that reason with
Boolean terms. In early work, Kotelnikov et al.\ 
\cite{kotelnikov-15-fool} have described a \emph{FOOL paramodulation}
rule that, under some order requirements, removes the need for the axiom
describing the Boolean domain---$\iforall p. \; p \ieq \itrue \llor p
\ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
arguments of symbols is translated to first-order logic.

@@

\looseness=-1 An important feature that we inherit by building on top of \lsup{}
is support for function extensionality. Moving to higher-order logic with
Booleans also means that we need to consider \emph{Boolean extensionality}:
$\iforall pq. \, (p \iequiv q) \iimplies p \ieq q$. We extend the rules of
\lsup{} that treat function extensionality to also treat Boolean
extensionality.

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the higher-order provers
Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
elimination of Leibniz equality, primitive instantiation, and treatment of choice
operator \cite{pa-01-classical-ty-thy}. We have also designed new rules that use
higher-order unification to resolve Boolean formulas that are hoisted to literal
level, delay clausification of nonatomic literals, reason about formulas under
$\lambda$-binders, and many others. Even though the rules we use are inspired by
the ones of refutationally complete higher-order provers, we do not guarantee
completeness of our extension of \lsup{}. Using the insights that we gained with
this incomplete extension of \lsup{} to full higher-order logic, Bentkamp
designed a complete calculus for full higher-order logic---\osup{}
\cite{bbtv-21-full-ho-sup}.

@@

Our approach is implemented in the Zipperposition theorem prover
\cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
to implement \lsup{}. We further extend
their implementation.

\looseness=-1
We performed an extensive evaluation of our approach (Sect.~\ref{sect:bool:eval}).
In addition to evaluating different configurations of our new rules, we have
compared them to full higher-order provers CVC4, Leo-III,  Satallax and Vampire.
The results suggest that it is beneficial to natively support Boolean reasoning---the approach outperforms preprocessing-based approaches. Furthermore, it is
very competitive with state-of-the-art higher order provers. We discuss the differences between our approach and the
approaches we base on, as well as related approaches (Sect.~
\ref{sect:bool:discussion}).

@@

We base our work (and parts of the following text) on Bentkamp et al.'s
\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. We
extend the syntax of this logic by adding logical connectives to the signature.
The semantic of the logic is extended by interpreting Boolean type $o$ as a
two-element domain. This amounts to extending the logic to full higher-order
logic. Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
types, and native logical connectives. For reference, we provide a
definition of this polymorphic logic.

@@

\looseness=-1
Function symbols $\cst{a}, \cst{b}, \cst{f}, \cst{g}, \ldots$ are elements of
$\Sigma$; their type declarations are written as $\cst{f} :
\forallty{\tuplen{\alpha}} \tau$. Set $\VV$ contains both free and bound variables, following
the distinction introduced in Sect.~\ref{sec:pre:hol}. Free term variables from the set $\VV$ are written
$F,G,X,Y, \ldots$ and we denote their types as $X : \tau$. Bound term variables are written $x,y,z,\ldots$,
and their types are similarly denoted. When the type is not
important, we omit type declarations. We assume that symbols $\itrue,
\ifalse,\inot,\iand, \ior,\iimplies,\iequiv$ with their standard meanings and type declarations are elements of
$\Sigma$. Furthermore, we assume that polymorphic symbols $\iforall$ and $\iexists$
with type declarations $\forallty{\alpha} (\alpha \rightarrow o) \rightarrow o$
and $\ieq \; : \forallty{\alpha} {\alpha \rightarrow \alpha \rightarrow o}$ are
in $\Sigma$, with their standard meanings. All these symbols are called \emph{logical
symbols}. We use infix notation for binary logical symbols.

@@

\looseness=-1
Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \ior
\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is a
nested Boolean, as well as its subterms. First-order logic allows only top-level
Booleans, whereas nested Booleans are characteristic for higher-order logic. 
Note that while our calculus works with clausal representation of the problem,
its higher-order nature allows representing formula $f$ as a singleton clause
containing only literal $f$.

@@

\looseness=-1
Zipperposition already had some support for Boolean reasoning before we
started extending \lsup{}. In this section, we first
describe the internals of Zipperposition responsible for reasoning with
Booleans. We continue by describing the rules that we have
implemented. For ease of presentation we divide them in three categories.
\subsection{Support for Booleans in Zipperposition}
\label{subsect:bool:zip-bools}

@@

In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
This means that higher-order term $(\cst{f} \, \cst{a}) \, cst{b}$ is
represented the same as first-order term $\cst{f}(\cst{a}, \cst{b})$. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
support for $\lambda$-terms is used to represent quantified nested Booleans:
Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
the input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

@@

\looseness=-1
Kotelnikov et al.\ \cite{kotelnikov-15-fool}, to the best of our
knowledge, pioneered the approach of extending a first-order superposition prover to support
nested Booleans. They call including the axiom $\iforall
p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
explosive behavior of the axiom, they imposed the following two requirements to
the simplification order $\succ$: $\itrue \succ \ifalse$ and
$\itrue$ and $\ifalse$ are two smallest ground terms with respect to $\succ$. If
these requirements are met, there is no self-paramodulation of the clause
and only paramodulation possible is from literal $p \eq \itrue$ of the mentioned axiom
into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
the axiom with the inference rule \emph{FOOL paramodulation} (\infname{FP}):
\pagebreak[2]
%
$$ \namedinference{FP}{C[f]}{C[\itrue] \vee f \eq \ifalse} $$
%
\looseness=-1
where $f$ is a nested non-variable Boolean subterm of clause $C$, different from
$\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
Booleans  to first-order logic without interpreted Booleans; thus, the symbols $\itrue$ and $\ifalse$, and the type $o$
correspond to proxy symbols and types introduced during the translation.

We created two rules that are syntactically similar to \infname{FP} but
are adapted for higher-order logic with one key distinction---we do not perform any translation:
%
\begin{align*}
  & \namedinference{Cases}{C[f]}{C[\ifalse] \vee f \eq \itrue}
  && \namedsimp{CasesSimp}{C[f]}{C[\ifalse] \vee f \eq \itrue \quad C[\itrue] \vee f \noteq \itrue}
\end{align*}
%
\looseness=-1
The prover that uses the rules should not include them both at the same time.
In addition, since
literals $f \eq \ifalse$ are represented as negative equations $f \noteq \itrue$, which cannot be used to paramodulate from,
we change the first requirement on the order to $\ifalse \succ
\itrue$, making $\itrue$ the smallest symbol.

\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
These two rules hoist Boolean subterms $f$ to the literal level; therefore,
some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
called immediate clausification (\infname{IC}), where $\tuple{D}{m}$ are defined below:
%
$$ \namedsimp{IC}{C}{D_1 \; \cdots \; D_m} $$
%
We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
is applicable if clause $C = L_1 \llor
\cdots \llor L_n$ is not standard.
The resulting clauses $\tuple{D}{m}$ represent
the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
Using Boolean extensionality, Zipperposition's clausification
algorithm treats Boolean equality as equivalence (i.e., it replaces
$\mathop{\eq}\typeargs{o}$ with $\iequiv$).

@@

\looseness=-1
If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result
in many redundant clauses. Consider the following example: let $\cst{p} : o
\rightarrow o$, $\cst{a} : o$ and consider a clause set containing $\cst{p} \,
(\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq \itrue$. Then, the clause $C
= \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in
eight ways using the rules, depending on which nested Boolean subterm was chosen
for the inference. In general, if a clause has a subterm occurrence of the form
$\cst{p}^n \, \cst{a}$, where both $\cst{p}$ and $\cst{a}$ have result type $o$,
the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
derived in at least $2^{n-1}$ ways.
% Boolean subterms occurring in different
% eligible literals cause similar explosive behavior.
To combat these issues we
implemented pragmatic restrictions of the rule: only $f$ which is
the leftmost outermost (or innermost) eligible subterm will be considered. With
this modification $C$ can be derived in only one way. Furthermore,
some intermediate conclusions of the rules will not be derived, pruning the search space.

The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
eagerly simplifies the input problem using Boolean equivalences before clausifying it. For example,
the formula $\cst p\iand\itrue$ is replaced by $\cst p$. To simplify nested Booleans we implemented the rule \infname{BoolSimp}
parameterized by a set of rewrite rules E:

\[
\namedsimp{BoolSimp}{C[\sigmaterm{f}]}{C[\sigmaterm{g}]}
\]
where $f \longrightarrow g \in E$ and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
described by Nonnengart and Weidenbach \cite[Sect.~3]{nw-01-small-cnf}. This set contains
the rules describing how each logical symbol behaves when given arguments $\itrue$ or $\ifalse$: for example, 
it includes $(\itrue \iimplies p)~\longrightarrow~p $ and $(p \iimplies \itrue) \longrightarrow \itrue$. %full list visa report Sect 4.2
Leo-III implements a similar rule, called \textsf{simp} \cite[Sect.~4.2.1.]{as-18-phd}.

@@

\subsection{Higher-Order Considerations}
\label{subsect:bool:core}
\looseness=-1
To achieve refutational completeness of higher-order resolution and similar
calculi, it is necessary to instantiate variables with result type $o$,
\emph{predicate variables}, with arbitrary formulas
\cite{as-18-phd,pa-01-classical-ty-thy}. Fortunately, we can approximate the
formulas using a complete set of logical symbols (e.g., $\inot$, $\iforall$, and
$\iand$). Since such an approximation is not only necessary for completeness of
some calculi, but possibly useful in practice, we implemented the \emph{primitive
instantiation} (PI) rule:
%
$$ \namedinference{PI} {C \llor (\lam{\tuple{x}{}}{P \, \tuplen{s}}) \, \dot{\eq}
\, t} {\substcl{\{ P
\mapsto f \}}{C \llor (\lam{\tuple{x}{}}{P \, \tuplen{s}}) \, \dot{\eq} \, t}  } $$
%
where $P$ is a free variable of
the type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow o$. 
Choosing a different $f$ that instantiates $P$, we can balance between
explosiveness of approximating a complete set of logical symbols and
incompleteness of pragmatic approaches. We borrow the notion of imitation from
higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}), and we say
that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
\, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $Z$ of type $\nu_1
\rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
are fresh free variables, where each $Y_i$ has the type $\nu_1 \rightarrow
\cdots \rightarrow \nu_m \rightarrow \tau_i$; variable $x_i$ is of type $\nu_i$.
\pagebreak[2]

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. Its implementation contains the following modes that generate
sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$. In  $\mathit{Imit}_{\star}$,
$\star$ is an element of a set of logical constants $S = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
generates imitations (for $P$) of all elements of $S$. Mode $\emph{Pragmatic}$
generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\ineq j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
\tau_j = o$, then it generates $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
generates $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ generates
imitations of $\itrue$, $\ifalse$ and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ which generates imitations
of both $\iforall$ and $\iexists$.

\looseness=-1
While experimenting with our implementation we noticed some proof patterns that
led us to come up with the following modifications. First, it often suffices to
perform \infname{PI} only on initial clauses---which is why we allow the rule to
be applied only to the clauses created using at most $k$ generating inferences.
Second, if the rule was used in the proof, its premise is usually only used as
part of that inference---which is why we implemented a version of \infname{PI}
that removes the clause after all possible \infname{PI} inferences have been
performed. We observed that the mode \emph{Imit}$_\star$ is useful in practice
since often approximation of a single logical symbol suffices.
